---
title: Analysis of OCT decays
author: Pascal PERNOT
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2
citecolor: blue
linkcolor: blue
linestretch: 1.1
fontsize: 12pt
bibliography: /home/pernot/Bureau/ComputChemUnc/biblio.bib
---

```{r setup, echo = FALSE, include=FALSE, message=FALSE}
libs =c('parallel','rstan','inlmisc','FitOCTLib')

for (lib in libs ) {
  if(!require(lib,character.only = TRUE))
    install.packages(lib,dependencies=TRUE)
  library(lib,character.only = TRUE)
}

# Options ####
options(mc.cores = parallel::detectCores(), width = 90)
rstan::rstan_options(auto_write = TRUE)
set.seed(1234) # Initialise la graine du RNG

# Graphical parameters ####
gPars = list(
  cols    = inlmisc::GetColors(8),
  col_tr  = inlmisc::GetColors(8,alpha=0.1), # Light, for spaghetti curves
  col_tr2 = inlmisc::GetColors(8,alpha=0.4), # Darker, for legends...
  pty     = 's',
  mar     = c(3,3,1.6,.2),
  mgp     = c(2,.75,0),
  tcl     = -0.5,
  lwd     = 1,
  cex     = 1,
  cex.leg = 0.8,
  xlabel  = 'stromal depth (µm)',
  plot_title = '',
  graphTable = FALSE
)

# Control parameters ####

### Default values / Set values here
ctrlPars = list(
  depthSel    = NULL, # Otherwise c(xmin,xmax)
  dataType    = 2,    # Intensity
  subSample   = 1,
  smooth_df   = 15,
  method      = c('sample','optim','vb')[1],
  nb_warmup   = 100,
  nb_sample   = 100,
  modRange    = 0.5,
  ru_theta    = 0.05,
  lambda_rate = 0.1,
  gridType    = 'internal',
  Nn          = 10,
  rho_scale   = 0.,
  priPost     = TRUE, # Compare prior and posterior pdf ?
  priorType   = 'abc'
)

# Expose parameters
for (n in names(ctrlPars))
  assign(n,rlist::list.extract(ctrlPars,n))

```


# Examples

## OCT signal

```{r getData1, echo = FALSE, cache = TRUE}
dataDir = c("DataWl","Data1","DataSynth")[2]
dataSet = (list.dirs(path=paste0('../',dataDir),full.names = FALSE)[-1])[1]

D = read.csv(paste0('../',dataDir,'/',dataSet,'/Courbe.csv'))
C = FitOCTLib::selX(D[,1],D[,2],depthSel,subSample)
x = C$x; y = C$y
```

### Estimate noise

A fit of the residuals of  smoothing splines provides the parameters for a model of the random noise.

```{r estimNoise1, echo = FALSE, results='hold'}
### Estimate data uncertainty
fits = FitOCTLib::estimateNoise(x, y, df = smooth_df)
uy   = fits$uy      # Used by next stages
ySpl = fits$ySmooth # Used by plotMonoExp

cat('\n Noise fit parameters:\n')
a = fits$theta
for(i in 1:length(a))
  cat(paste0('a_',i,' : '),signif(a[i],3),'\n')
```

The smoothing curve, residuals and noise model can be seen in Fig. \@ref(fig:plotNoise1). The noise presents a notable exponential decay.

```{r plotNoise1, echo = FALSE, fig.cap='Splines smoothing and noise estimation'}
gPars$plot_title = 'Noise estimation'
FitOCTLib::plotNoise(x, y, uy, ySpl, gPars, dataType)
gPars$plot_title = ''
```

### Mono-exponential fit

A mono-exponential fit is attempted using the noise model.

```{r fitMono1, include =FALSE}
fitm   = FitOCTLib::fitMonoExp(x, y, uy, dataType = dataType)
theta0    = fitm$best.theta   # Used by next stage
cor.theta = fitm$cor.theta
```

The Birge ratio analysis rejects this model

```{r fitMonoOut1, echo = FALSE, results='hold'}
opt = theta0
cat('\n MonoExp decay parameters:\n')
for(i in 1:length(opt))
  cat(paste0('b_',i,' : '),signif(opt[i],3),'\n')
cat('\n\n')

# Probability Interval for Birge's ratio
br = FitOCTLib::printBr(fitm$fit)
```

It can be seen in Fig. \@ref(fig:plotMono1) that the residuals present notable serial correlation.

```{r plotMono1, echo = FALSE, fig.cap='Mono-exponential fit and residuals.'}
fit     = fitm$fit
resid   = fit$par$resid
mod     = fit$par$m
gPars$plot_title = 'Mono-exponential fit'
FitOCTLib::plotMonoExp(x, y, uy, ySpl, mod, resid,
                       gPars, dataType, br)
gPars$plot_title = ''
```

### Modulated-exponential fit

Considering that the mono-exponential is deemed inadequate, the
model with mean-depts modulation is fit to the data.
The default parameters are used, notably with a number of control
points $M=10$. 

#### Prior sampling

```{r priGP1, include=FALSE, cache=TRUE, dependson='getData1'}
# Prior on exponential params
priExp = FitOCTLib::estimateExpPrior(
  x, uy, dataType, priorType,
  out = fitm, ru_theta = ru_theta,
  eps = 1e-3
)

# Posterior Distribution
priGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 1,
  open_progress = FALSE
)
```


The moments-matching prior is used for the exponential parameters,
and one can check the dispersion of the exponential curves generated 
from this prior (prior predictive sampling) in Fig. \@ref(fig:plotPriGP1).
Prior probability intervals for the control values are also shown in
this figure (right panel).

```{r plotPriGP1, echo = FALSE, fig.cap='Samples from prior pdf.'}
gPars$plot_title = 'Prior exponential'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=priGP, modScale=modRange*4, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```

#### Posterior sampling

The priors pdf seems correct, and one can now draw samples from 
the posterior pdf.

```{r fitGP1, include=FALSE, cache=TRUE, dependson='getData1'}
# Posterior Distribution
fitGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 0,
  open_progress = FALSE
)
```

The Birge ratio analysis validates this model.

```{r fitGPOut1, echo = FALSE, results ='hold', message= FALSE}
if(fitGP$prior_PD == 0) {
  # cat('\n ExpGP parameters:\n')
  fit = fitGP$fit
  # if(fitGP$method == 'sample' | fitGP$method == 'vb') {
  #   pars = c('theta','yGP','lambda','sigma','br')
  #   print(fit,pars=pars)
  #   br = mean(rstan::extract(fit,'br')[[1]])
  # } else {
  #   cat('theta  :',fit$par$theta,'\n')
  #   cat('yGP    :',fit$par$yGP,'\n')
  #   cat('lambda :',fit$par$lambda,'\n')
  #   cat('sigma  :',fit$par$sigma,'\n')
  #   cat('br     :',br<-fit$par$br,'\n')
  # }
  # cat('\n\n')
  
  # Probability Interval for Birge's ratio
  br = FitOCTLib::printBr(fit)
}
```

As can be seen in Fig. \@ref(fig:plotGP1), the residuals have been
notably improved from the mono-exponential model.

```{r plotGP1, echo = FALSE, fig.height=7, fig.cap='Modulated-exponential fit, residuals, Modulation function and Normal QQ-plot of residuals.'}
gPars$cex = 1
gPars$plot_title = 'Modulated exp. fit'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=fitGP, modScale=modRange, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```


## In-vivo signal

```{r getData, echo = FALSE, cache = TRUE}
dataDir = c("DataWl","Data1","DataSynth")[2]
dataSet = (list.dirs(path=paste0('../',dataDir),full.names = FALSE)[-1])[4]

D = read.csv(paste0('../',dataDir,'/',dataSet,'/Courbe.csv'))
    C = FitOCTLib::selX(D[,1],D[,2],depthSel,subSample)
    x = C$x; y = C$y
```

### Estimate noise

A fit of the residuals of  smoothing splines provides the parameters for a model of the random noise.

```{r estimNoise, echo = FALSE, results='hold'}
### Estimate data uncertainty
fits = FitOCTLib::estimateNoise(x, y, df = smooth_df)
uy   = fits$uy      # Used by next stages
ySpl = fits$ySmooth # Used by plotMonoExp

cat('\n Noise fit parameters:\n')
a = fits$theta
for(i in 1:length(a))
  cat(paste0('a_',i,' : '),signif(a[i],3),'\n')
```

The smoothing curve, residuals and noise model can be seen in Fig. \@ref(fig:plotNoise). In thsi example, the noise is homogeneous, 
and the exponential model tends to a constant.

```{r plotNoise, echo = FALSE, fig.cap='Splines smoothing and noise estimation'}
gPars$plot_title = 'Noise estimation'
FitOCTLib::plotNoise(x, y, uy, ySpl, gPars, dataType)
gPars$plot_title = ''
```

### Mono-exponential fit

A mono-exponential fit is attempted using the noise model.

```{r fitMono, include =FALSE}
fitm   = FitOCTLib::fitMonoExp(x, y, uy, dataType = dataType)
theta0    = fitm$best.theta   # Used by next stage
cor.theta = fitm$cor.theta
```

The Birge ratio analysis rejects this model

```{r fitMonoOut, echo = FALSE, results='hold'}
opt = theta0
cat('\n MonoExp decay parameters:\n')
for(i in 1:length(opt))
  cat(paste0('b_',i,' : '),signif(opt[i],3),'\n')
cat('\n\n')

# Probability Interval for Birge's ratio
br = FitOCTLib::printBr(fitm$fit)
```

It can be seen in Fig. \@ref(fig:plotMono) that the residuals present notable serial correlation.

```{r plotMono, echo = FALSE, fig.cap='Mono-exponential fit and residuals.'}
fit     = fitm$fit
resid   = fit$par$resid
mod     = fit$par$m
gPars$plot_title = 'Mono-exponential fit'
FitOCTLib::plotMonoExp(x, y, uy, ySpl, mod, resid,
                       gPars, dataType, br)
gPars$plot_title = ''
```

### Modulated-exponential fit

Considering that the mono-exponential is deemed inadequate, the
model with mean-depts modulation is fit to the data.
The default parameters are used, notably with a number of control
points $M=10$. 

#### Prior sampling

```{r priGP, include=FALSE, cache=TRUE, dependson='getData'}
# Prior on exponential params
priExp = FitOCTLib::estimateExpPrior(
  x, uy, dataType, priorType,
  out = fitm, ru_theta = ru_theta,
  eps = 1e-3
)

# Posterior Distribution
priGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 1,
  open_progress = FALSE
)
```


The moments-matching prior is used for the exponential parameters,
and one can check the dispersion of the exponential curves generated 
from this prior (prior predictive sampling) in Fig. \@ref(fig:plotPriGP).
Prior probability intervals for the control values are also shown in
this figure (right panel).

```{r plotPriGP, echo = FALSE, fig.cap='Samples from prior pdf.'}
gPars$plot_title = 'Prior exponential'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=priGP, modScale=modRange*4, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```

#### Posterior sampling

The priors pdf seems correct, and one can now draw samples from 
the posterior pdf.

```{r fitGP, include=FALSE, cache=TRUE, dependson='getData'}
# Posterior Distribution
fitGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 0,
  open_progress = FALSE
)
```

The Birge ratio analysis invalidates this model.

```{r fitGPOut, echo = FALSE, results ='hold', message= FALSE}
if(fitGP$prior_PD == 0) {
  # cat('\n ExpGP parameters:\n')
  fit = fitGP$fit
  # if(fitGP$method == 'sample' | fitGP$method == 'vb') {
  #   pars = c('theta','yGP','lambda','sigma','br')
  #   print(fit,pars=pars)
  #   br = mean(rstan::extract(fit,'br')[[1]])
  # } else {
  #   cat('theta  :',fit$par$theta,'\n')
  #   cat('yGP    :',fit$par$yGP,'\n')
  #   cat('lambda :',fit$par$lambda,'\n')
  #   cat('sigma  :',fit$par$sigma,'\n')
  #   cat('br     :',br<-fit$par$br,'\n')
  # }
  # cat('\n\n')
  
  # Probability Interval for Birge's ratio
  br = FitOCTLib::printBr(fit)
}
```

As can be seen in Fig. \@ref(fig:plotGP), the residuals have not been
notably improved from the mono-exponential model, and the present design
does not enable to fit a sharp rise in the OCT signal at a depth of about
$400\, \mu\rm{m}$. An option is to increase the number of control points.

```{r plotGP, echo = FALSE, fig.height=7, fig.cap='Modulated-exponential fit, residuals, Modulation function and Normal QQ-plot of residuals.'}
gPars$cex = 1
gPars$plot_title = 'Modulated exp. fit'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=fitGP, modScale=modRange, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```

### Add control points

```{r include=FALSE}
Nn = 15
gridType='extremal'
```


```{r fitGP2, include=FALSE, cache=TRUE, dependson='getData'}
# Posterior Distribution
fitGP2 = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 0,
  open_progress = FALSE
)
```

The number of control points has been raised to _M_=`r Nn` 
on a grid covering the extremities of the depts range, 
and the posterior distribution has been resampled. 

```{r fitGPOut2, echo = FALSE, results ='hold', message= FALSE}
if(fitGP2$prior_PD == 0) {
  cat('\n ExpGP parameters:\n')
  fit = fitGP2$fit
  pars = c('theta','yGP','lambda','sigma','br')
  print(fit,pars=pars)
  cat('\n\n')
  # Probability Interval for Birge's ratio
  br = FitOCTLib::printBr(fit)
}
```

The Birge ratio is now acceptable, although there are only 3 
active control points. 
The others do not contribute significantly to the fit.
The residuals are also improved (Fig. \@ref(fig:plotGP2)).
The model is statistically valid, and the modulation curve can be used 
for physical interpretation.

```{r plotGP2, echo = FALSE, fig.height=7, fig.cap='Modulated-exponential fit, residuals, Modulation function and Normal QQ-plot of residuals.'}
gPars$cex = 1
gPars$plot_title = 'Modulated exp. fit'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=fitGP2, modScale=modRange, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```





# References

