---
title: Analysis of OCT decays
author: Pascal PERNOT
date: 2018/12/03
output: 
  bookdown::pdf_document2
citecolor: blue
linkcolor: blue
linestretch: 1.1
fontsize: 11pt
bibliography: /home/pernot/Bureau/ComputChemUnc/biblio.bib
---

```{r setup, echo = FALSE, include=FALSE, message=FALSE}
libs =c('parallel','rstan','inlmisc','FitOCTLib')

for (lib in libs ) {
  if(!require(lib,character.only = TRUE))
    install.packages(lib,dependencies=TRUE)
  library(lib,character.only = TRUE)
}

# Options ####
options(mc.cores = parallel::detectCores(), width = 90)
rstan_options(auto_write = TRUE)
set.seed(1234) # Initialise la graine du RNG

# Graphical parameters ####
gPars = list(
  cols    = inlmisc::GetColors(8),
  col_tr  = inlmisc::GetColors(8,alpha=0.1), # Light, for spaghetti curves
  col_tr2 = inlmisc::GetColors(8,alpha=0.4), # Darker, for legends...
  pty     = 's',
  mar     = c(3,3,1.6,.2),
  mgp     = c(2,.75,0),
  tcl     = -0.5,
  lwd     = 1,
  cex     = 1,
  xlabel  = 'stromal depth (µm)',
  plot_title = '',
  graphTable = FALSE
)

# Control parameters ####

### Default values / Set values here
ctrlPars = list(
  depthSel    = NULL, # Otherwise c(xmin,xmax)
  dataType    = 2,    # Intensity
  subSample   = 1,
  smooth_df   = 15,
  method      = c('sample','optim','vb')[1],
  nb_warmup   = 100,
  nb_sample   = 100,
  modRange    = 0.5,
  ru_theta    = 0.05,
  lambda_rate = 0.1,
  gridType    = 'internal',
  Nn          = 10,
  rho_scale   = 0.1,
  priPost     = TRUE, # Compare prior and posterior pdf ?
  priorType   = 'abc'
)

# Expose parameters
for (n in names(ctrlPars))
  assign(n,rlist::list.extract(ctrlPars,n))

```



# Introduction

Analysis of the OCT decays by a mono-exponential decay reveals two
features which condition our design of a data analysis method:
an heterogeneous random noise, and medium-scale oscillations 
around the exponential decay (model inadequacy). 
To be able to partition unambiguously the model residuals between 
these two components, we proceed in two steps:

1. estimation and modeling of the random noise component, to be injected in 

2. the estimation of the parameters of the decay model:
    a. test of a simple monoexponential model
    b. if the latter fails, use of a modulated decay model


# Methods

Considering a set of $N$ measured data points $\pmb D=\{z_i,y_i \}_{i=1, N}$,
one considers a measurement model with heterogeneous additive noise
\begin{equation}  
  y_i = f(z_i) + \epsilon_i 
(\#eq:1)
\end{equation}    
where $f(.)$ is a model function to be defined, and
\begin{equation}  
  \epsilon_i \sim Norm(0,\sigma_i) 
(\#eq:2)
\end{equation}  
represents an heterogeneous measurement noise with a normal distribution
of standard deviation $\sigma_i$.

## Estimation and modeling of the random noise

A cubic smoothing spline function is used to estimate the random 
part of the signal. 
The residuals of the smoothing function are assigned 
to random noise $\epsilon$.

Considering that the OCT decays result from photon counting experiments,
one can expect that the noise obbeys a Poisson law. 
In consequence, the standard deviation of the noise is modeled by an exponential decay
\begin{equation} 
  \sigma_i = a_1 * \exp\left( -\frac{2*z_i}{a_2} \right) 
(\#eq:3)
\end{equation}  
This shape enables also to account for cases of _in vivo_ measurements
with nearly uniform noise. 

The parameters are obtained by Bayesian inference [@Gelman2013] from 
Eqns. 2-3, with uniform priors in the range $]0,a_{max}]$, the upper 
value being chosen to accommodate a quasi-uniform noise model. 

The posterior pdf is sampled by Markov Chain Monte Carlo in `stan`
[@Gelman2015] (see below for details of implementation), and the mean
values are used to define the measurement uncertainty
\begin{equation} 
  u_{yi} = \bar{a_1} * \exp\left( -\frac{2*z_i}{\bar{a_2}} \right) 
(\#eq:4)
\end{equation}  
to be used in the next steps, for which one has now a data set 
augmented with measurement uncertainties $\pmb u_y$, _i.e._ 
$\pmb D= \{ z_i,y_i, u_{yi} \}_{i=1, N}$.

## Calibration of a decay model

### Mono-exponential decay

The mono-exponential decay curve with parameters 
$\pmb \vartheta = \{ a,b,l_0 \}$
\begin{equation} 
  f( z; \pmb \vartheta) = a + b * \exp\left( -\frac{2*z}{l_0} \right)
(\#eq:modExp)
\end{equation}  
is first fitted to the data by maximization of the posteriot pdf (MAP).
The likelihood is
\begin{equation}
  \pmb y| \pmb \vartheta \sim \exp\left\{ -\frac{1}{2}\ 
        \chi_w^2(\pmb y; \pmb z, \pmb u_y, \pmb \vartheta) \right\}
(\#eq:likExp)
\end{equation}
where the weighted chi-square function is defined as
\begin{equation}
  \chi_w^2(\pmb y; \pmb z, \pmb u_y, \pmb \vartheta) = 
    \sum_{i=1}^N 
      \frac{[y_i - f(z_i; \pmb \vartheta )]^2}
           {u_{yi}^2}
(\#eq:wChi2)
\end{equation}
The parameters have uniform priors on $[0,\infty[$.

#### Validation

The value of the reduced chi-square $\chi_r^2 = \chi_w^2/(N-3)$ should be 
close to 1, based on the quantiles of the reduced chi-square distribution 
with $N-3$ degrees of freedom. 
Moreover, the residuals should not present serial correlation. 
If these conditions are not met, one has to use a more elaborate model,
described below.


### Modulated decay model

The mono-exponential decay model is improved with a $z$-dependent 
optical depth $l(z; \pmb \vartheta)$ with parameters $\pmb \vartheta$
\begin{equation} 
  f( z; \pmb \vartheta, \pmb \kappa) = a + b * 
         \exp\left( -\frac{2*z}{l(z; l_0,\pmb \kappa)} \right)
(\#eq:5)
\end{equation}  
where the shape of the optical depth is defined as
\begin{equation} 
  l(z; l_0,\pmb \kappa) = l_0*(1+ \delta l(z; \pmb \kappa))
(\#eq:6)
\end{equation}
$\delta l(.)$ is a Gaussian Process (GP) of mean 0, 
conditioned on $M$ control values 
$\pmb \kappa= \{\kappa_i\}_{i=1,M}$
at predefined locations 
$\hat{\pmb z} = \{\hat z_i\}_{i=1,M}$.
The mean value of the GP is used here as an interpolator between 
the control points, and we choose a Gaussian kernel for its 
smoothness properties
\begin{equation} 
C(z, z') = \alpha^2 * \exp\left( -\frac{(z-z')^2}{\rho^2} \right)
(\#eq:7)
\end{equation}
The $\alpha$ and $\rho$ parameters of the GP are fixed *a priori*.

Considering the set of $M$ control values $\pmb \kappa$ for the OD 
modulation at locations $\hat{\pmb z}$, $\delta l$ can be obtained 
at any depth as the mean value of the GP:
\begin{equation} 
  \delta l(z; \pmb \kappa) = 
     {\pmb \Omega}^T*{\pmb K}^{-1}*{\pmb \kappa}
(\#eq:8)
\end{equation}
where ${\pmb K}$ is a $M\times M$ covariance matrix with elements 
$K_{ij} = C(\hat z_i, \hat z_j )$ and ${\pmb \Omega}$ is a $M$-vector 
with elements $\Omega_i = C(\hat z_i, z )$.

The control points positions, $\hat{\pmb z}$, are chosen *a priori* 
on a regular grid spanning the experimental depth range. 
As one does not expect short scale modulations, a small number of 
points is used, typically $M\simeq 10$. 
Similarly, for the smoothness of the interpolation, one picks 
the correlation length of the kernel *a priori*, at a value large 
enough to avoid undue oscillations between the control points and 
small enough to avoid excessive rigidity of the model. 
In the present configuration, a good compromise has been found 
to be $\rho = 1/M^{th}$ of the total depth range.
In the same spirit, the variance parameter of the GP is taken as a small
fraction of the standard deviation of the control values 
$\alpha = 0.1 * sd(\pmb \kappa)$. 
This choice of $\alpha$ and $\rho$ has been found to provide a well
behaved interpolator for test simulated signals. 
Besides, small changes around these values do not affect 
significantly the mean prediction of the GP (Eq. 8).

#### Likelihood function

To compensate for defaults in the estimation of the noise, 
a parameter $\sigma$ is introduced as a multiplicative factor
of $\pmb u(y)$.
The unknown parameters are therefore $\pmb \vartheta$, $\pmb \kappa$,
and $\sigma$.
The likelihood function is a multivariate Normal distribution
\begin{equation}
    \pmb y | \pmb \vartheta, \pmb \kappa , \sigma  \sim 
      \exp\left\{ -\frac{1}{2} \chi_w^2(\pmb y; \pmb z, \pmb u_y,
      \pmb \vartheta, \pmb \kappa, \sigma) \right\}
(\#eq:likMod)
\end{equation}
where
\begin{equation}
  \chi_w^2(\pmb y; \pmb z, \pmb u_y, \pmb \vartheta, \pmb \kappa, \sigma) = 
    \sum_{i=1}^N
    \frac{[y_i - f(z_i; \pmb \vartheta, \pmb \kappa )]^2}
         {[\sigma * u_{yi}]^2}
(\#eq:wChi2Mod)
\end{equation}

#### Prior pdfs

There is a risk of indetermination between $l_0$ and $\pmb \kappa$; 
for instance, setting all values of $\pmb \kappa$ to 1 would be 
exactly compensated by halving $l_0$.

One therefore constrains $\pmb \kappa$ to be close to zero
with a Bayesian Lasso-type prior [@Park2008], in a version 
based on a hierarchical prior adapted from Ref.[@Mallick2014]:
\begin{equation}
  \begin{split}
    \kappa_i | u_i & \sim Normal(0, s_i); i=1,M \\
    s_i | \lambda & \sim  Gamma(2, \lambda) ; i=1,M\\
    \lambda & \sim Gamma(2, \lambda_r)
  \end{split}
(\#eq:9)
\end{equation}

The prior on $\pmb \vartheta$ is a multivariate normal distribution
$\pi(\pmb\vartheta) = N(\hat{\pmb \vartheta}_1,\pmb \Sigma_\vartheta)$
centered on the best estimate from the mono-exponential fit, 
$\hat{\pmb \vartheta}_1$, with a covariance matrix 
$\pmb \Sigma_\vartheta$ estimated following two options:

1. the covariance matrix is built from the correlation matrix
   $\pmb C_{\vartheta_1}$ issued from the monoexponential fit, and a 
   vector of standard deviations specified from relative 
   uncertainties on the parameters:
   \begin{equation}
      \pmb \Sigma_\vartheta = \pmb I(\pmb u_\vartheta) 
                       * \pmb C_{\vartheta_1} 
                       * \pmb I(\pmb u_\vartheta)
   (\#eq:sig1)
   \end{equation}
   where $\pmb I(\pmb u_\vartheta)$ is a diagonal matrix with elements
   $\pmb u_\vartheta = r * \hat{\pmb \vartheta}_1$.
   The uncertainty factor $r$ is typically chosen as a small percentage,
   _e.g._, $r = 0.05$.
   
2. a _diagonal_ covariance matrix 
   $\pmb \Sigma_\vartheta = \pmb I(\pmb u_\vartheta^2)$
   is built from a moments-matching procedure.
   The standard deviations $\pmb u_\vartheta$ are optimized to 
   match two criteria:  
    a. $S_1$, the 2-sigma prediction uncertainty of the mono-exponential
       model has to match $Q_{95}$, the $95^{th}$ quantile of the absolute 
       errors of the mono-exponential model (all statistics are weighted 
       by $\pmb u_y$) [@Pernot2018];   
    b. the standard deviation of the prediction uncertainty has to 
       be as small as possible.  
    The first criterion ensures that the mean prediction uncertainty of
    the mono-exponential model is in agreement with the amplitude of
    the model's residuals. This criterion can typically be matched by 
    an infinity of solutions, and the second one selects those parameters
    which provides the 'flatest' prediction band. 
    The optimization is done by sampling the posterior pdf of the
    parameters  $\pmb u_\vartheta$ with uniform priors on $[0,\infty[$,
    and a likelihood function
    \begin{equation}
      S_1 ,S_2 | \pmb u_\vartheta \sim 
        \exp\left\{ 
          -\frac{1}{2}\frac{(S_1 - Q_{95})^2 + S_2^2}
                           {\varepsilon^2} 
          \right\}
    (\#eq:likMM)
    \end{equation}
    where $S_1 = 1.96*\sqrt{<\pmb u_p^2/\pmb u_y^2>}$, $\pmb u_p$ is the
    prediction uncertainty of the mono-exponential model estimated 
    by linear uncertainty propagation [@GUM], 
    $Q_{95}$ is the the $95^{th}$ quantile of the absolute weighted 
    residuals
    $|\left\{\pmb y - f(\pmb z;\pmb \vartheta)\right\} / \pmb u_y|$,
    $S_2 = {\rm sd}(\pmb u_p / \pmb u_y)$
    and $\varepsilon$ is a predefined precision factor.
  
The prior on $\sigma$ is a normal distribution, centered on 1, 
with standard deviation 0.1.

#### Validation

The quality of the fit can be estimated by inspection of
the residuals which should not present serial correlations 
and should conform with the random experimental noise. 

The value of the reduced chi-square $\chi_r^2 = \chi_w^2/(N-\nu)$, 
where $\nu$ is the number of effective free parameters, should be 
close to 1. One has $\nu = 5 + \hat M$, where $\hat M$ is the number of 
control values significantly different from zero, _i.e._ those
for which 0 lies outside of a high posterior probability interval.
One typically considers a 0.9 probability interval.

Posterior predictive samples are also generated and plotted 
with the reference data to confirm the quality of the fit.

# Implementation

__Algorithm__

1. Noise estimation
    i. Perform a smoothing of data and extract residuals $\pmb R$  
    ii. Fit residuals by a normal distribution with exponentially 
       decreasing amplitude

2. Mono-exponential fit

3. Modulated exponential fit


Estimation of the random errors is done with the `smooth.spline` from 
`R` [@RCoreTeam]; A satisfying degree of smoothing for all examples
considered here was obtained by setting the smoothing `df` parameter 
to 15.

The Bayesian models are implemented in `stan` [@Gelman2015], using the
`rstan` interface package [@Rstan] for `R` [@RCoreTeam]. 
`Stan` is a very flexible and efficient probabilistic programming 
language to implement Bayesian statistical models. 
The No-U-Turn sampler [@Hoffman2014] was used for this study.

The main outputs of the stan codes are samples of the posterior pdf of
the parameters, from which statistics and plots can be generated in R.
Convergence of the sampling is assessed by examining the traces of
parameters samples and the 'split Rhat' statistics provided by rstan. In
the present application, the Markov chains converge rapidly, and all
models are run with four parallel Markov Chains of 1500 iterations each,
1000 of which are used as warm-up for the No-U-Turn sampler and
dispatched. The convergence criteria and parameters statistics are
therefore estimated on a sample of 2000 points.

# Example(s)

## In-vivo signal

```{r getData, echo = FALSE, cache = TRUE}
dataDir = c("DataWl","Data1","DataSynth")[2]
dataSet = (list.dirs(path=paste0('../',dataDir),full.names = FALSE)[-1])[4]

D = read.csv(paste0('../',dataDir,'/',dataSet,'/Courbe.csv'))
    C = FitOCTLib::selX(D[,1],D[,2],depthSel,subSample)
    x = C$x; y = C$y
```

## Estimate noise

A fit of the residuals of  smoothing splines provides the parameters for a model of the random noise.

```{r estimNoise, echo = FALSE, results='hold'}
### Estimate data uncertainty
fits = FitOCTLib::estimateNoise(x, y, df = smooth_df)
uy   = fits$uy      # Used by next stages
ySpl = fits$ySmooth # Used by plotMonoExp

cat('\n Noise fit parameters:\n')
a = fits$theta
for(i in 1:length(a))
  cat(paste0('a_',i,' : '),signif(a[i],3),'\n')
```

The smoothing curve, residuals and noise model can be seen in Fig. \@ref(fig:plotNoise). In thsi example, the noise is homogeneous, 
and the exponential model tends to a constant.

```{r plotNoise, echo = FALSE, fig.cap='Splines smoothing and noise estimation'}
gPars$plot_title = 'Noise estimation'
FitOCTLib::plotNoise(x, y, uy, ySpl, gPars, dataType)
gPars$plot_title = ''
```

## Mono-exponential fit

A mono-exponential fit is attempted using the noise model.

```{r fitMono, include =FALSE}
fitm   = FitOCTLib::fitMonoExp(x, y, uy, dataType = dataType)
theta0    = fitm$best.theta   # Used by next stage
cor.theta = fitm$cor.theta
```

The Birge ratio analysis rejects this model

```{r fitMonoOut, echo = FALSE, results='hold'}
opt = theta0
cat('\n MonoExp decay parameters:\n')
for(i in 1:length(opt))
  cat(paste0('b_',i,' : '),signif(opt[i],3),'\n')
cat('\n\n')

# Probability Interval for Birge's ratio
br = FitOCTLib::printBr(fitm$fit)
```

It can be seen in Fig. \@ref(fig:plotMono) that the residuals present notable serial correlation.

```{r plotMono, echo = FALSE, fig.cap='Mono-exponential fit and residuals.'}
fit     = fitm$fit
resid   = fit$par$resid
mod     = fit$par$m
gPars$plot_title = 'Mono-exponential fit'
FitOCTLib::plotMonoExp(x, y, uy, ySpl, mod, resid,
                       gPars, dataType, br)
gPars$plot_title = ''
```

## Modulated-exponential fit

Considering that the mono-exponential is deemed inadequate, the
model with mean-depts modulation is fit to the data.
The default parameters are used, notably with a number of control
points $M=10$. 

### Prior sampling

```{r priGP, include=FALSE, cache=TRUE, dependson='getData'}
# Prior on exponential params
priExp = FitOCTLib::estimateExpPrior(
  x, uy, dataType, priorType,
  out = fitm, ru_theta = ru_theta,
  eps = 1e-3
)

# Posterior Distribution
priGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 1,
  open_progress = FALSE
)
```


The moments-matching prior is used for the exponential parameters,
and one can check the dispersion of the exponential curves generated 
from this prior (prior predictive sampling) in Fig. \@ref(fig:plotPriGP).
Prior probability intervals for the control values are also shown in
this figure (right panel).

```{r plotPriGP, echo = FALSE, fig.cap='Samples from prior pdf.'}
gPars$plot_title = 'Prior exponential'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=priGP, modScale=modRange*4, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```

### Posterior sampling

The priors pdf seems correct, and one can now draw samples from 
the posterior pdf.

```{r fitGP, include=FALSE, cache=TRUE, dependson='getData'}
# Posterior Distribution
fitGP = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 0,
  open_progress = FALSE
)
```

The Birge ratio analysis invalidates this model.

```{r fitGPOut, echo = FALSE, results ='hold', message= FALSE}
if(fitGP$prior_PD == 0) {
  # cat('\n ExpGP parameters:\n')
  fit = fitGP$fit
  # if(fitGP$method == 'sample' | fitGP$method == 'vb') {
  #   pars = c('theta','yGP','lambda','sigma','br')
  #   print(fit,pars=pars)
  #   br = mean(rstan::extract(fit,'br')[[1]])
  # } else {
  #   cat('theta  :',fit$par$theta,'\n')
  #   cat('yGP    :',fit$par$yGP,'\n')
  #   cat('lambda :',fit$par$lambda,'\n')
  #   cat('sigma  :',fit$par$sigma,'\n')
  #   cat('br     :',br<-fit$par$br,'\n')
  # }
  # cat('\n\n')
  
  # Probability Interval for Birge's ratio
  br = FitOCTLib::printBr(fit)
}
```

As can be seen in Fig. \@ref(fig:plotGP), the residuals have not been
notably improved from the mono-exponential model, and the present design
does not enable to fit a sharp rise in the OCT signal at a depth of about
$400\, \mu\rm{m}$. An option is to increase the number of control points.

```{r plotGP, echo = FALSE, fig.height=7, fig.cap='Modulated-exponential fit, residuals, Modulation function and Normal QQ-plot of residuals.'}
gPars$cex = 1
gPars$plot_title = 'Modulated exp. fit'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=fitGP, modScale=modRange, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```

### Add control points

```{r include=FALSE}
Nn = 15
```


```{r fitGP2, include=FALSE, cache=TRUE, dependson='getData'}
# Posterior Distribution
fitGP2 = FitOCTLib::fitExpGP(
  x, y, uy,
  dataType      = dataType,
  Nn            = Nn,
  gridType      = gridType,
  method        = method,
  theta0        = priExp$theta0,
  Sigma0        = priExp$Sigma0,
  lambda_rate   = lambda_rate,
  rho_scale     = ifelse(rho_scale==0, 1./Nn, rho_scale),
  nb_warmup     = nb_warmup,
  nb_iter       = nb_warmup + nb_sample,
  prior_PD      = 0,
  open_progress = FALSE
)
```

The number of control points has been raised to $M = `r Nn` $, 
and the posterior distribution has been resampled. 
The Birge ratio is now acceptable

```{r fitGPOut2, echo = FALSE, results ='hold', message= FALSE}
if(fitGP2$prior_PD == 0) {
  # cat('\n ExpGP parameters:\n')
  fit = fitGP2$fit
  # if(fitGP$method == 'sample' | fitGP$method == 'vb') {
  #   pars = c('theta','yGP','lambda','sigma','br')
  #   print(fit,pars=pars)
  #   br = mean(rstan::extract(fit,'br')[[1]])
  # } else {
  #   cat('theta  :',fit$par$theta,'\n')
  #   cat('yGP    :',fit$par$yGP,'\n')
  #   cat('lambda :',fit$par$lambda,'\n')
  #   cat('sigma  :',fit$par$sigma,'\n')
  #   cat('br     :',br<-fit$par$br,'\n')
  # }
  # cat('\n\n')
  
  # Probability Interval for Birge's ratio
  br = FitOCTLib::printBr(fit)
}
```

and the residuals are considerably improved (Fig. \@ref(fig:plotGP2)).
The model is statistically valid, and the modulation curve can be used 
for physical interpretation.

```{r plotGP2, echo = FALSE, fig.height=7, fig.cap='Modulated-exponential fit, residuals, Modulation function and Normal QQ-plot of residuals.'}
gPars$cex = 1
gPars$plot_title = 'Modulated exp. fit'
FitOCTLib::plotExpGP(
  x, y, uy, ySpl, out=fitGP2, modScale=modRange, gPars=gPars,
  dataType = dataType, br = br
)
gPars$plot_title = ' '
```





# References

